```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(patchwork)

# Read the data
expanded_data <- read_csv("data/expanded_filtered_merged_data.csv")

```

# NA value

## Check NA Values
```{r}
# Check for NA values in each column
na_summary <- expanded_data %>%
  summarise(across(everything(), ~ mean(is.na(.)) * 100, .names = "NA_percent_{.col}"))

# Preview NA summary
print(na_summary)
```

##  Filter Columns with NA Proportion > 50%
```{r}
# Filter out columns where NA proportion is greater than 50%
filtered_data <- expanded_data %>%
  select(where(~ mean(is.na(.)) <= 0.5))

str(filtered_data)
```

## Strategy for Imputation

1.	Linear Interpolation (Time-Series Trend):  
- Interpolate missing values for each variable (across years) within each country based on its existing data.  
2.	Fallback Imputation:  
- If a country has NA for all years for a variable, impute using the global mean or median of that variable.
```{r}
# Function to handle imputation per country
impute_country_data <- function(filtered_data) {
  filtered_data %>%
    mutate(across(where(is.numeric), ~ na.approx(., na.rm = FALSE))) # Linear interpolation
}

# Step 1: Apply interpolation for each country
imputed_data <- filtered_data %>%
  group_by(country) %>%
  group_modify(~ impute_country_data(.)) %>%
  ungroup()

# Step 2: Global mean/median imputation for columns with remaining NA
imputed_data <- imputed_data %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

# Check for remaining NA values
na_summary_after <- imputed_data %>%
  summarise(across(everything(), ~ sum(is.na(.)), .names = "NA_count_{.col}"))

# Save the final imputed data
write_csv(imputed_data, "data/imputed_data.csv")
```

# EDA

## Summary Statistics for Each Column
```{r}
# Compute basic summary statistics
summary_stats <- imputed_data %>%
  summarise(across(where(is.numeric), list(
    mean = ~mean(., na.rm = TRUE),
    median = ~median(., na.rm = TRUE),
    sd = ~sd(., na.rm = TRUE),
    min = ~min(., na.rm = TRUE),
    max = ~max(., na.rm = TRUE)
  )))

# Save summary statistics for review
write_csv(summary_stats, "data/summary_statistics_imputed.csv")

# Display summary statistics
print(summary_stats)
```

## Correlation Matrix for 2020 Data
```{r}
# Filter data for the year 2020
data_2020 <- imputed_data %>%
  select(country, countryiso3code, contains("_2020"))

# Compute correlation matrix for numeric variables
cor_matrix <- data_2020 %>%
  select(-country, -countryiso3code) %>% # Exclude non-numeric columns
  cor(use = "complete.obs")

# Convert correlation matrix to long format for plotting
cor_data <- as.data.frame(as.table(cor_matrix))

cor_plot <- ggplot(cor_data, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0, limit = c(-1, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Correlation Matrix (2020 Data)", fill = "Correlation")

cor_plot
# Save the plot
ggsave("plot/correlation_matrix_2020_imputed.png", plot = cor_plot, width = 10, height = 8)
```

```{r}
# Filter numeric columns and calculate correlation with life_expectancy_2020
cor_with_target <- data_2020 %>%
  select(-country, -countryiso3code) %>%  # Exclude non-numeric columns
  summarise(across(everything(), ~ cor(.x, life_expectancy_2020, use = "complete.obs"))) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "correlation")

# Filter variables with correlation larger than 0.7
strong_correlations <- cor_with_target %>%
  filter(abs(correlation) > 0.5)

# Display the strongly correlated variables
print(strong_correlations)
```


## Visualize Distributions for 2020 Variables
```{r}
# Pivot data to long format for visualization
long_data_2020 <- data_2020 %>%
  pivot_longer(cols = -c(country, countryiso3code), names_to = "variable", values_to = "value")

# Plot distributions of variables
distribution_plot <- ggplot(long_data_2020, aes(x = value)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  theme_minimal() +
  labs(title = "Distributions of Variables (2020 Data)", x = "Value", y = "Frequency")
distribution_plot
# Save the plot
ggsave("data/distribution_2020_imputed.png", plot = distribution_plot, width = 12, height = 10)
```

## Explore Relationships Between Variables
```{r}
# Variables to plot
variables_to_plot <- c("gdp_per_capita_2020", "undernourishment_rate_2020", 
                       "health_expenditure_2020", "infant_mortality_rate_2020")

# Create individual scatterplots
plot1 <- ggplot(data_2020, aes(x = gdp_per_capita_2020, y = life_expectancy_2020)) +
  geom_point(alpha = 0.7, color = "blue") +
  theme_minimal() +
  labs(title = "GDP per Capita vs Life Expectancy",
       x = "GDP per Capita (2020)", y = "Life Expectancy (2020)")

plot2 <- ggplot(data_2020, aes(x = undernourishment_rate_2020, y = life_expectancy_2020)) +
  geom_point(alpha = 0.7, color = "red") +
  theme_minimal() +
  labs(title = "Undernourishment Rate vs Life Expectancy",
       x = "Undernourishment Rate (2020)", y = "Life Expectancy (2020)")

plot3 <- ggplot(data_2020, aes(x = health_expenditure_2020, y = life_expectancy_2020)) +
  geom_point(alpha = 0.7, color = "green") +
  theme_minimal() +
  labs(title = "Health Expenditure vs Life Expectancy",
       x = "Health Expenditure (2020)", y = "Life Expectancy (2020)")

plot4 <- ggplot(data_2020, aes(x = infant_mortality_rate_2020, y = life_expectancy_2020)) +
  geom_point(alpha = 0.7, color = "purple") +
  theme_minimal() +
  labs(title = "Infant Mortality Rate vs Life Expectancy",
       x = "Infant Mortality Rate (2020)", y = "Life Expectancy (2020)")

# Combine plots into a 2x2 grid
combined_plot <- (plot1 + plot2) / (plot3 + plot4)
combined_plot
# Save the combined plot
ggsave("plot/relationships_life_expectancy_2020.png", plot = combined_plot, width = 16, height = 10)


# Standardize the data
data_2020 <- data_2020 %>%
  mutate(
    gdp_per_capita_2020 = scale(gdp_per_capita_2020),
    undernourishment_rate_2020 = scale(undernourishment_rate_2020),
    health_expenditure_2020 = scale(health_expenditure_2020),
    infant_mortality_rate_2020 = scale(infant_mortality_rate_2020)
  )

# Reshape the data to long format

data_long <- data_2020 %>%
  pivot_longer(
    cols = c(gdp_per_capita_2020, undernourishment_rate_2020, health_expenditure_2020, infant_mortality_rate_2020),
    names_to = "variable",
    values_to = "value"
  ) %>%
  mutate(variable = factor(variable, levels = c(
    "gdp_per_capita_2020", "undernourishment_rate_2020", "health_expenditure_2020", "infant_mortality_rate_2020"
  )))

# Create a static plot with multiple regression lines and custom legend labels

static_plot <- ggplot(data_long, aes(x = value, y = life_expectancy_2020, color = variable)) +
  geom_point(alpha = 0.7) +
  geom_smooth(se = FALSE) +
  theme_minimal() +
  labs(
    title = "Standardized Variables vs Life Expectancy (2020)",
    x = "Standardized Variable Value",
    y = "Life Expectancy (2020)",
    color = "Variable"
  ) +
  scale_color_discrete(labels = c(
    "gdp_per_capita_2020" = "GDP per Capita",
    "undernourishment_rate_2020" = "Undernourishment Rate",
    "health_expenditure_2020" = "Health Expenditure",
    "infant_mortality_rate_2020" = "Infant Mortality Rate"
  ))
static_plot
# Convert the static plot to an interactive plot using ggplotly


interactive_plot <- ggplotly(static_plot, tooltip = c("x", "y", "color")) %>%
  layout(legend = list(title = "Variable"),
         xaxis = list(title = "Standardized Variable Value"),
         yaxis = list(title = "Life Expectancy (2020)"),
         legend = list(traceorder = "normal",
                       itemsizing = "constant",
                       itemclick = "toggle",
                       itemdoubleclick = "toggleothers"))

# Manually set the legend labels in ggplotly

for (i in 1:length(interactive_plot$x$data)) {
  interactive_plot$x$data[[i]]$name <- c(
    "GDP per Capita",
    "Undernourishment Rate",
    "Health Expenditure",
    "Infant Mortality Rate"
  )[i]
}

# Display the interactive plot

interactive_plot

# UI

ui <- fluidPage(
  plotlyOutput("plot")
)

# Server

server <- function(input, output) {
  output$plot <- renderPlotly({
    interactive_plot

  })
}

# Run the application

shinyApp(ui = ui, server = server)

htmlwidgets::saveWidget(interactive_plot, "interactive_plot.html")
```


